Using an iterated block particle filter via \pkg{spatPomp}

### 2022-12-15 ms.Rnw
full recompute
tidying

%%%%%% iiiiiiii %%%%
<<bm-model,echo=T,eval=T>>=
library(spatPomp)
i <- 1
b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
plot(b)
@

@@

<<ibpf-prof-rho-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

@@

abline(v=p1bpf_mcap$ci)

@@

abline(v=p1kf_mcap$ci)
abline(v=p1exact_mcap$ci,col="red",lty="dashed")

@@

abline(v=p1pf_mcap$ci)

@@

<<ibpf-prof-sig-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

@@

abline(v=p1_sig_bpf_mcap$ci)

@@

abline(v=p1_sig_kf_mcap$ci)
abline(v=p1_sig_exact_mcap$ci,col="red",lty="dashed")

@@

abline(v=p1_sig_pf_mcap$ci)

@@

<<m-prof-R0-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

@@

abline(v=m_p2_R0_bpf_mcap$ci)

@@

abline(v=m_p2_R0_pf_mcap$ci)

@@

<<m_slice_g_plot,eval=T,echo=F,fig.height=4, fig.width=6, out.width="5.5in", fig.cap = "Slice for $g_4$ through the true parameter vector, using BPF evaluation with $K=4$ blocks each having 1 unit.">>=

@@

abline(v=m_slice_g_mcap$ci)

@@

The block approximation in BPF concerns dependence between blocks and therefore may have an effect on estimation of parameters describing the coupling between units.
The measles metapopulation model has a so-called gravity model for coupling, with a parameter $g_u$ controlling the rate of transmission from other cities into city $u$.
A relatively straightforward way to investigate estimation of $g_u$ using BPF and IBPF is to compute a likelihood slice through the true parameter value for a simulation, with only $g_u$ being varied.
Here, we investigate a slice for $u=4$.
A likelihood profile cannot take a lower value than a likelihood slice, since the profile has an additional optimization.
Therefore, a flat slice implies a flat profile; the converse is not necessarily true.
Clear evidence of bias in a slice for a small number of units would anticipate difficulties when undertaking the more time-consuming task of obtaining a profile with many units and many parameters.
Figure~\ref{fig:m_slice_g_plot} suggests that the BPF likelihood approximation leads to a low-bias estimate of $g_u$ when compared with Figures~\ref{fig:ibpf-prof-rho-plot} and~\ref{fig:ibpf-prof-sig-plot}.
CHECK THAT THIS MAKES SENSE IN THE CONTEXT OF THE FINAL RESULTS.

>>>>

%%%%%% iiiiiiii %%%%
<<bm-model,echo=T,eval=T>>=
library(spatPomp)
i <- 2
b <- bm2(U=4,N=switch(i,10,200),unit_specific_names="rho")
plot(b)
@

@@

<<ibpf-prof-rho-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter. Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.",eval=T,echo=F>>=

@@

abline(v=p1bpf_mcap$ci)
abline(v=p1bpf_mcap$mle)
abline(v=coef(b)["rho1"],lty="dotted")

@@

abline(v=p1kf_mcap$ci)
abline(v=p1exact_mcap$ci,col="red",lty="dashed")
abline(v=p1kf_mcap$mle)
abline(v=coef(b)["rho1"],lty="dotted")

@@

abline(v=p1pf_mcap$ci)
abline(v=p1pf_mcap$mle)
abline(v=coef(b)["rho1"],lty="dotted")

@@

<<ibpf-prof-sig-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter. Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.",eval=T,echo=F>>=

@@

abline(v=p1_sig_bpf_mcap$ci)
abline(v=p1_sig_bpf_mcap$mle)
abline(v=coef(b_sig)["sigma1"],lty="dotted")

@@

abline(v=p1_sig_kf_mcap$ci)
abline(v=p1_sig_exact_mcap$ci,col="red",lty="dashed")
abline(v=p1_sig_kf_mcap$mle)
abline(v=coef(b_sig)["sigma1"],lty="dotted")

@@

abline(v=p1_sig_pf_mcap$ci)
abline(v=p1_sig_pf_mcap$mle)
abline(v=coef(b_sig)["sigma1"],lty="dotted")

@@

<<m-prof-R0-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $R0_1$ using an IBPF search with likelihood computed using BPF, for $K=4$ blocks each having 1 unit. Bottom: The same IBPF search with likelihood computed using the particle filter. Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.",eval=T,echo=F>>=

@@

abline(v=m_p2_R0_bpf_mcap$ci)
abline(v=m_p2_R0_bpf_mcap$mle)
abline(v=coef(m_R0)[paste0("R0",m_R0_prof_u)],lty="dotted")

@@

abline(v=m_p2_R0_pf_mcap$ci)
abline(v=m_p2_R0_pf_mcap$mle)
abline(v=coef(m_R0)[paste0("R0",m_R0_prof_u)],lty="dotted")

@@

<<m_slice_g_plot,eval=T,echo=F,fig.height=4, fig.width=6, out.width="5.5in", fig.cap = "Slice for $g_4$ through the true parameter vector, using BPF evaluation with $K=4$ blocks each having 1 unit. Vertical lines show the MLE and a 95\\% confidence interval, with a dotted line at the true parameter value.">>=

@@

abline(v=m_slice_g_mcap$ci)
abline(v=m_slice_g_mcap$mle)
abline(v=coef(m_g)[paste0("g",m_g_prof_u)],lty="dotted")

@@

The block approximation in BPF concerns dependence between blocks and therefore may have an effect on estimation of parameters describing the coupling between units.
The measles metapopulation model has a so-called gravity model for coupling, with a parameter $g_u$ controlling the rate of transmission from other cities into city $u$.
A relatively straightforward way to investigate estimation of $g_u$ using BPF and IBPF is to compute a likelihood slice through the true parameter value for a simulation, with only $g_u$ being varied.
Here, we investigate a slice for $u=4$.
A likelihood profile cannot take a lower value than a likelihood slice, since the profile has an additional optimization.
Therefore, a flat slice implies a flat profile; the converse is not necessarily true.
Clear evidence of bias in a slice for a small number of units would anticipate difficulties when undertaking the more time-consuming task of obtaining a profile with many units and many parameters.
Figure~\ref{fig:m_slice_g_plot} suggests that the BPF likelihood approximation leads to a lower-bias estimate of $g_u$ than for the coupling parameter $\rho$ in the Gaussian model (Figure~\ref{fig:ibpf-prof-rho-plot}).
Figure~\ref{fig:m_slice_g_plot} is consistent with a small positive bias: BPF ignores part of the dependence in the filter distribution, so it may be expected that it compensates by a bias toward parameter values that increase the coupling.
